Extracting The Power of Dependent Types

### 2021-09-23 paper.tex
Addressing reviewer's comments (hopefully).

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Ad\-ding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-con\-struc\-tion programs in languages without built-in verification
  capabilities, while maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest.  Thirdly, we
  extract a program written in the original language,
  so that it can be used by the existing compilers and tools.

@@

There are two common approaches to language embedding: deep and shallow.
A shallow embedding defines a language in terms of functions and operators
of the host language.  Programs in a shallowly-embedded language are programs
in the host language, so they are immediately run\-nable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of embedded programs.

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\citep{PasalicTS02}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types, tagless
interpreters are technically
possible~\citep{CHAPMAN200921,10.1007/978-3-540-74464-1_7}, but are
extremely challenging to define and use in practice.

@@

To keep presentation as easy as possible to follow, we have
chosen a simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
While the embedding may look simplistic, this simplicity
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

This entire paper is a literate Agda script: all code is typechecked
while generating the paper.

@@

@@

%\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

>>>>

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Ad\-ding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-con\-struc\-tion programs in languages without built-in verification
  capabilities, while maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest within the embedding.  Thirdly, we
  extract a program written in the original language,
  so that it can be used by the existing compilers and tools.

@@

There are two common approaches to language embedding: deep and shallow.
Shallow embeddings define languages in terms of functions and operators
of the host language.  Programs in shallowly-embedded languages are programs
in the host language, so they are immediately run\-nable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of embedded programs.

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\citep{PasalicTS02,Tagless}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types, tagless
interpreters are technically
possible~\citep{CHAPMAN200921,10.1007/978-3-540-74464-1_7}, but are
\emph{extremely challenging} to define and use in practice.
For further details on why exactly this is challenging refer to the
first section of~\cite{10.1145/1863495.1863497} where the problem is meticulously
explained by~\citeauthor{10.1145/1863495.1863497}.

@@

To keep presentation as easy as possible to follow, we have
chosen a simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
This paper is not concerned with practical application of
the presented embedding, so we consciously keep the setup as minimal as
possible.  It is powerful enough to motivate the use of dependent
types \emph{within} the embedding, and to demonstrate the essence
of the proposed approach.  While we include many patterns found
in bigger languages, we leave the scaling of the approach out
of the scope of the paper referring the interested reader
to~\cite{DBLP:journals/corr/abs-2105-10819} where we work
out such generalisations.

% While the embedding may look simplistic, this simplicity
% allows us to focus on the essence of the proposed approach and avoid unrelated
% technical details.  We ensure that many verification patterns found in bigger
% languages are covered, showing that the proposed approach scales.

This entire paper is a literate Agda script\footnote{The sources of the paper
can be found at \url{https://github.com/ashinkarov/agda-stacklang}.}:
all code is typechecked while generating the paper.

@@

When dealing with embedded languages that are extracted and run
by external compilers, we are fundamentally bound to have potential
semantic mismatches.  Semantics of the embedding may differ from the
semantics implemented by external compilers, as well as translation from the
embedding back to external compilers can be faulty.  Real world compilers
face exactly the same problem: how do you know that generated assembly
code is doing what you think it is doing?  Addressing such problems
requires fully verified toolchains that run on a verified hardware.
Enormous research efforts are spent towards addressing this
problem, but this is beyond the scope of this paper.

@@

%\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This
means that our proposed approach is immediately implementable
in these languages --- one simply needs to adjust extraction
to the particular API.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.
